Follow-up Samples

### 2017-02-23 submission/manuscript_outline_20161024.Rmd
read through after all additions

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e. adenoma or carcinoma) removal.  Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.

@@

**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal cancer lesion removal.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models.  The first model was built to classify lesion and normal (non-lesion) samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma or carcinoma removal.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.  Finally, a list of common OTUs were found for the two different models used.

@@

@@

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e. adenoma and carcinoma) removal.  Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.

@@

**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal lesion removal.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma and carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models.  The first model was built to classify lesion versus non-lesion (normal) samples while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial versus follow up.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after removal [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after lesion removal [Figure 2].

@@

Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma removal.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after adenoma and carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.  Finally, a list of common OTUs were found for the two different models used.

@@

The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.

@@

Not applicable.
